# Software user guide

This section describes how to install and run the model.

## Installing the package from Github


### Windows users
Before you install the `FIMS` package, you will need to install the Rtools executable corresponding to your `R` version as well as the `TMB` package and its dependencies. `FIMS` has only been developed and tested on `R` version 4.+, and so in order to install the package you will need to ensure you are using `R` version 4.+ and an RStudio version that is at least 1.2.5042. Instructions on how to install `Rtools` are [here](https://cran.r-project.org/bin/windows/Rtools/rtools40.html). Instructions on how to install `TMB` are [here](https://github.com/kaskr/adcomp/wiki/Download). Please ensure you have tested your `TMB` setup before moving on to install FIMS.

### Mac users

### Linux users

## Installing from `R`

```{r eval=FALSE}
remotes::install_github("NOAA-FIMS/FIMS")
library(FIMS)
```

## Specifying and compiling the model
You can add components to the model using S4 classes.
```{r}
#TODO: add script to populate the model
```


## Extracting model output

Here is how you get the model output.

```{r}
#Todo add code for how to extract model output
```

