# Software developer guide

This section describes the software you will need to contribute to this project. This is in addition to the software dependencies described in the [software user guide](#software-user-guide) which you should ensure are installed first.

## git

You will need git installed locally, and you may prefer to use an additional git [GUI client](https://git-scm.com/downloads/guis) such as GitKraken or Github Desktop. If your preferred git client is the RStudio IDE, you can configure Git and RStudio integration following these [instructions](https://happygitwithr.com/rstudio-git-github.html). 

To install git, please follow the instructions on [this page](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) for your operating system. You can find the downloads for your operating system on the left-hand navigation bar on that page.

## Development environment

An integrated development environment is recommended to organize code files, outputs, and build and debug environments. The most popular IDEs on the development team are [RStudio](https://www.rstudio.com/) and [Visual Studio Code](https://code.visualstudio.com/). You are welcome to use another IDE if you'd like. 

## C++ compiler
Windows users who installed Rtools should have a C++ compiler (`gcc`) as part of the bundle. To ensure the C++ compiler is on your path, open a command prompt and type `gcc`. If you get the below message, you are all set:

```{R eval = FALSE}
gcc: fatal error: no input files
compilation terminated.
```

If not, you will need to check that the compiler is on the path. The easiest way to do so is by creating a text file .Renviron in your Documents folder which contains the following line:
```
PATH="${RTOOLS40_HOME}\usr\bin;${PATH}"
```

You can do this with a text editor, or from R like so (note that in R code you need to escape backslashes):
```{r eval=FALSE}
write('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', file = "~/.Renviron", append = TRUE)
```

Restart R, and verify that make can be found, which should show the path to your Rtools installation.

```{r eval=FALSE}
Sys.which("make")
## "C:\\rtools40\\usr\\bin\\make.exe"
```

## Google test

You will need to install `CMake` and `ninja` and validate you have the correct setup by following the steps outlined in the [test case template](#test-case-template).
