# User guide

This section details installation guides for [users](#user-installation-guide). See the [developer installation guide](#developer-installation-guide).

## User Installation Guide
This section describes how to install the FIMS R package and dependencies.

### Installing the package from Github


#### Windows users
Before you install the `FIMS` package, you will need to install the Rtools executable corresponding to your `R` version as well as the `TMB` package and its dependencies. `FIMS` has only been developed and tested on `R` version 4.+, and so in order to install the package you will need to ensure you are using `R` version 4.+ and an RStudio version that is at least 1.2.5042. Instructions on how to install `Rtools` are [here](https://cran.r-project.org/bin/windows/Rtools/rtools40.html). Instructions on how to install `TMB` are [here](https://github.com/kaskr/adcomp/wiki/Download). Please ensure you have tested your `TMB` setup before moving on to install FIMS.

#### Mac users

#### Linux users

### Installing from `R`

```{r eval=FALSE}
remotes::install_github("NOAA-FIMS/FIMS")
library(FIMS)
```

## Running the model

This section describes how to set-up and run the model.

### Specifying the model

#### Naming conventions
TODO: add description and link to naming conventions

#### Structuing data input
You can add components to the model using S4 classes.
```{r}
#TODO: add script to demonstrate how to structure data input
```

#### Defining model specifications
```{r}
#TODO: add scripts detailing how to set up different components of the model
```

### How to run the model
```{r}
#TODO: add script with examples on how to run the model
```

### Extracting model output

Here is how you get the model output.

```{r}
#Todo add code for how to extract model output
```


