# User guide

This section details installation guides for [users](#user-installation-guide). See the [developer installation guide](#developer-installation-guide).

## User Installation Guide
This section describes how to install the FIMS R package and dependencies.

## Installing the package from Github

The following software is required:
- R version 4.0.0 or newer (or RStudio version 1.2.5042 or newer)
  - the `remotes` R package
- `TMB` (install instructions at are
[here](https://github.com/kaskr/adcomp/wiki/Download).)

### Windows users

- Rtools4 (available from [here](https://cran.r-project.org/bin/windows/Rtools/rtools40.html))
  - this likely requires IT support to install it on NOAA computers
  (or any without administrative accounts)

## Installing from `R`

```{r eval=FALSE}
remotes::install_github("NOAA-FIMS/FIMS")
library(FIMS)
```

## Running the model

This section describes how to set-up and run the model.

### Specifying the model

#### Naming conventions
TODO: add description and link to naming conventions

#### Structuing data input
You can add components to the model using S4 classes.
```{r}
#TODO: add script to demonstrate how to structure data input
```

#### Defining model specifications
```{r}
#TODO: add scripts detailing how to set up different components of the model
```

### How to run the model
```{r}
#TODO: add script with examples on how to run the model
```

### Extracting model output

Here is how you get the model output.

```{r}
#Todo add code for how to extract model output
```


