# Developer Software and Installation Guide

This section describes the software you will
need to contribute to this project as well as ways FIMS can be installed.

## Codespaces as an alternative to installing software

An alternative to setting up this software locally is using [GitHub Codespaces](https://docs.github.com/en/codespaces).

The FIMS repository already has a file to set up all of the needed software, so there is no need to install anything locally.

To start a codespace on the FIMS repository, follow the [GitHub documentation for creating a Codespace](https://docs.github.com/en/codespaces/developing-in-a-codespace/creating-a-codespace-for-a-repository?tool=webui#creating-a-codespace-for-a-repository).

## Software to Install

### git

You will need git installed locally, and you may prefer to use an
additional git [GUI client](https://git-scm.com/downloads/guis) such as
GitKraken or GitHub Desktop. If your preferred git client is the RStudio
IDE, you can configure Git and RStudio integration following these
[instructions](https://happygitwithr.com/rstudio-git-GitHub.html).

To install git, please follow the instructions on [this
page](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) for
your operating system. You can find the downloads for your operating
system on the left-hand navigation bar on that page.

### Development environment

An integrated development environment is recommended to organize code 
files, outputs, and build and debug environments. The most popular IDEs
 on the development team are [RStudio](https://www.rstudio.com/) and 
 [Visual Studio Code](https://code.visualstudio.com/). You are welcome 
 to use another IDE or a text-editor based workflow if you strongly 
 prefer that. 

### vscode setup

Install [Visual Studio Code](https://code.visualstudio.com/Download) and
[configure it for using `R`](https://github.com/REditorSupport/vscode-R/wiki/Installation:-Windows).

For those migrating from R studio to VS code, [this post on migrating to VS
code](https://statnmap.com/2021-10-09-how-not-to-be-lost-with-vscode-when-coming-from-rstudio/)
may be helpful.

A number of optional settings that could be added to the user settings
(settings.json) file in vscode to improve the usability of R in VS code.
For example, the [settings for interacting with R
terminals](https://github.com/REditorSupport/vscode-R/wiki/Interacting-with-R-terminals#sending-code-to-r-terminals)
can be adjusted.

Here are some that you may want to use with FIMS:

```json
{
    // Associate .RMD files with markdown:
    "files.associations": {
        "*.Rmd": "markdown",
    },
    // A cmake setting
    "cmake.configureOnOpen": true, 
    // Set where the rulers are, needed for Rewrap. 72 is the default we have
    // decided on for FIMS repositories.z
    "editor.rulers": [
        72
    ],
    // Should the editor suggest inline edits?
    "editor.inlineSuggest.enabled": true,
    // Settings for github copilot and which languages to use it with or not.
    "github.copilot.enable": {
        "*": true,
        "yaml": false,
        "plaintext": false,
        "markdown": false,
        "latex": false,
        "r": false
    }, 
    // Setting for sending R code from the editor to the terminal
    "r.alwaysUseActiveTerminal": true,
    // Needed to send large chunks of code to the r terminal when using radian
    "r.bracketedPaste": true,
    // Needed to use httpgd for plotting in vscode
    "r.plot.useHttpgd": true,
    // path to the r terminal (in this case, radian). Necessary to get the terminal to use radian.
    "r.rterm.windows": "C://Users//my.name//AppData//Local//Programs//Python//Python310//Scripts//radian.exe", //Use this only for Windows 
    // options for the r terminal
    "r.rterm.option": [
        "--no-save",
        "--no-restore",
        "max.print=500"
    ],
    // Setting for whether to allow linting of documents or not
    "r.lsp.diagnostics": true,
    // When looking at diffs under the version control tab, should whitspace be ignored?
    "diffEditor.ignoreTrimWhitespace": false,
    // What is the max number of lines that are printed as output to the terminal?
    "terminal.integrated.scrollback": 10000
}
```

Some [suggested R
shortcuts](https://github.com/REditorSupport/vscode-R/wiki/Keyboard-shortcuts)
could be helpful.


To set up C++ with vscode, instructions are 
[here](https://code.visualstudio.com/docs/languages/cpp).

Other helpful extensions that can be found in the VScode marketplace are:
 - Github Copilot: An AI tool that helps with line completion
 - Live Share: Collaborate on the same file remotely with other developers
 - Rewrap: Helps rewrapping comments and text lines at a specified character
   count. Note that to get it working it will be necessary to add
   [rulers](https://code.visualstudio.com/docs/getstarted/tips-and-tricks#_vertical-rulers)
 - There are a number of keymap packages that import key mappings from commonly
  used text editors (e.g., Sublime, Notepad++, atom, etc.). Searching "keymap"
  in the marketplace should help find these.
  - GitLens (or GitLess): Adds more Git functionality. Note that some of the
    GitLens functionality is not free, and GitLess is a fork before the addition
    of these premium features.

Note that the keybindings.json and settings.json could be copied from one
computer to another to make it easier to set up VS code with the settings
needed. Note that the [settings.json location differs](https://code.visualstudio.com/docs/getstarted/settings#_settings-file-locations) depending on the operating system.

Typically, it is good practice to not restore old sessions after shutting down
the IDE. To avoid restoring old sessions in the VS Code terminals (including R
terminal), in the Setting User Interface within VS Code (get to this by opening
the command palette and searching for Preferences: Open Settings (UI)), under
Features > Terminal, uncheck the option "Enable Persistent Sessions."

Rstudio addins can be accessed by searching for Rstudio addin in the command
palette. Clicking on "R: Launch Rstudio Addin" should provide a list of addin
options.

### C++ compiler

Windows users who installed Rtools should have a C++ compiler (`gcc`) 
as part of the bundle. To ensure the C++ compiler is on your path, open
 a command prompt and type `gcc`. If you get the below message, you are
  all set:

```{R eval = FALSE} 
gcc: fatal error: no input files compilation
terminated.
```

If not, you will need to check that the compiler is on the path. The 
easiest way to do so is by creating a text file .Renviron in your 
Documents folder which contains the following line:
```
PATH="${RTOOLS44_HOME}\usr\bin;${PATH}"
```

You can do this with a text editor, or from `R` like so (note that in 
`R` code you need to escape backslashes):
```{r eval=FALSE}
write('PATH="${RTOOLS44_HOME}\\usr\\bin;${PATH}"', file = "~/.Renviron", append = TRUE)
```

Restart R, and verify that make can be found, which should show the path
to your Rtools installation.

```{r eval=FALSE}
Sys.which("make") ##
"C:\\rtools44\\usr\\bin\\make.exe" 
```

### GoogleTest

You will need to install `CMake` and `ninja` and validate you have the
correct setup by following the steps outlined in the [test case
template](#testing).

### GDB debugger

Windows users who use GoogleTest may need GDB debugger to see what is 
going on inside of the program while it executes, or what the program is 
doing at the moment it crashed. rtools44 includes the GDB debugger. The 
steps below help install 64-bit version gdb.exe. 

- Open Command Prompt and type `gdb`. If you see details of usage, GDB
debugger is already in your PATH. If not, follow the instructions below
to install GDB debugger and add it to your PATH. 
- Install Rtools run commands in the mingw64 shell. Run 
command `pacman -Sy mingw-w64-x86_64-gdb` to install 64-bit version 
(more information can be found in [R on Windows FAQ](https://github.com/r-windows/docs/blob/master/faq.md#does-rtools40-include-a-debugger))
- Type `Y` in the mingw64 shell to proceed with installation
- Check whether ~/rtools44/mingw64/bin/gdb.exe exists or not
- Add rtools44 to the PATH and you can check that the path is working 
by running `which gdb` in a command window

### Doxygen

To build C++ documentation website for FIMS, a documentation 
generator Doxygen needs to be installed. Doxygen automates the generation of 
documentation from source code comments. To install Doxygen, please follow the 
instructions [here](https://www.doxygen.nl/manual/install.html) to install 
Doxygen on various operating systems. Below are steps to install 64-bit version of
Doxygen 1.11.0 on Windows. 

- Download [doxygen-1.11.0.windows.x64.bin.zip](https://www.doxygen.nl/files/doxygen-1.11.0.windows.x64.bin.zip) 
and extract the applications to `Documents\Apps\Doxygen` or other preferred folder.

- Add Doxygen to the PATH by following similar instructions [here](https://noaa-fims.github.io/collaborative_workflow/testing.html#adding-cmake-and-ninja-to-your-path-on-windows).

- Open a command window and run `where doxygen` to check if Doxygen is
added to the PATH.

- Two commands on the command line are needed to generate C++ documentation 
for FIMS locally:

```bash
cmake -S. -B build -G Ninja
cmake -- build build
```

### R Installation Requirements

The following is needed:
- R version 4.0.0 or newer (or RStudio version 1.2.5042 or newer)
- the `devtools` R package
- `TMB` version 1.8.0 or newer ([install instructions](https://github.com/kaskr/adcomp/wiki/Download))
- For windows users: Rtools4 (available [on the CRAN website](https://cran.r-project.org/bin/windows/Rtools/rtools40.html))

## Installing FIMS

### Install Options for Developers

Use the following options from the R console:

- Clone FIMS and build using `devtools::load_all()` (mimics package building) or `devtools::install()` (actually builds package from local files). These are good options if you intend to modify files in FIMS and want to test them out.
- Use `install.packages("FIMS", repos = c("https://noaa-fims.r-universe.dev", "https://cloud.r-project.org"))` to download the latest precompiled version from R universe or `devtools::install_github("NOAA-FIMS/FIMS)` to compile yourself. `devtools::install_github()` has options to all you to be more specific about which version of the code is being built (e.g., you can refer to a tag or branch, which is not possible with the R universe install).

### Windows: Fixing Fatal Error

Windows users can expect to see some derivative of the following error message in their R session if they have not yet set some flags using {withr}.
```
Fatal error: can't write <xxx> bytes to section .text of FIMS.o: 'file too big
```
You can easily fix this by running the following line in your local R session. Note that this call will need to be repeated each time you open a new session.
```{r eval=FALSE}
withr::local_options(pkg.build_extra_flags = FALSE)
```
This fix does not work when `devtools::test()` is called and FIMS is re-compiled. The call to `devtools::test()` in this case overwrites the local options set by withr. Compile FIMS first using `devtools::load_all()` prior to running `devtools::test()`.

This fix removes the debugger flag `-O0 -g` from being automatically inserted for certain devtools calls (e.g. `devtools::load_all()`). Windows developers wanting to compile FIMS with the debugger turned on will need to run the above script in addition to manually modifying the call to PKG_CXXFLAGS in the [Makevars.win](https://github.com/NOAA-FIMS/FIMS/blob/doc-install/src/Makevars.win) file in the src directory to the following:

```
PKG_CXXFLAGS =  -DTMB_MODEL  -DTMB_EIGEN_DISABLE_WARNINGS -O1 -g
```
To turn off the debugger flag, remove the `-O1 -g` flag:
```
PKG_CXXFLAGS =  -DTMB_MODEL  -DTMB_EIGEN_DISABLE_WARNINGS
```

## Getting Help
Please report bugs along with a minimal reproducible example on GitHub [Issues](https://github.com/NOAA-FIMS/FIMS/issues)


