# Developer Software Guide

This section describes the software you will
need to contribute to this project. This is in addition to the software
dependencies described in the [user installation
guide](#user-installation-guide) which you should ensure are installed
first.

### git

You will need git installed locally, and you may prefer to use an
additional git [GUI client](https://git-scm.com/downloads/guis) such as
GitKraken or GitHub Desktop. If your preferred git client is the RStudio
IDE, you can configure Git and RStudio integration following these
[instructions](https://happygitwithr.com/rstudio-git-GitHub.html).

To install git, please follow the instructions on [this
page](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) for
your operating system. You can find the downloads for your operating
system on the left-hand navigation bar on that page.

### Development environment

An integrated development environment is recommended to organize code 
files, outputs, and build and debug environments. The most popular IDEs
 on the development team are [RStudio](https://www.rstudio.com/) and 
 [Visual Studio Code](https://code.visualstudio.com/). You are welcome 
 to use another IDE or a text-editor based workflow if you strongly 
 prefer that. 

 ### vscode setup

 Please follow the instructions [here](https://code.visualstudio.com/docs/languages/r) 
 to set up vscode for use with `R`.

 In addition to those instructions, you may need to install the `vscDebugger` package
 [here](https://github.com/ManuelHentschel/vscDebugger) using the command:
 ```{r eval=FALSE}
 remotes::install_github("ManuelHentschel/vscDebugger")
 ```

 To set up C++ with vscode, instructions are 
 [here](https://code.visualstudio.com/docs/languages/cpp).

 Other helpful extensions are Github Copilot and Live Share which you
  can find by searching in the marketplace.

### C++ compiler

Windows users who installed Rtools should have a C++ compiler (`gcc`) 
as part of the bundle. To ensure the C++ compiler is on your path, open
 a command prompt and type `gcc`. If you get the below message, you are
  all set:

```{R eval = FALSE} 
gcc: fatal error: no input files compilation
terminated.
```

If not, you will need to check that the compiler is on the path. The 
easiest way to do so is by creating a text file .Renviron in your 
Documents folder which contains the following line:
```
PATH="${RTOOLS40_HOME}\usr\bin;${PATH}"
```

You can do this with a text editor, or from `R` like so (note that in 
`R` code you need to escape backslashes):
```{r eval=FALSE}
write('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', file = "~/.Renviron", append = TRUE)
```

Restart R, and verify that make can be found, which should show the path
to your Rtools installation.

```{r eval=FALSE}
Sys.which("make") ##
"C:\\rtools40\\usr\\bin\\make.exe" 
```

### GoogleTest

You will need to install `CMake` and `ninja` and validate you have the
correct setup by following the steps outlined in the [test case
template](#testing).
