# Contributor Guidelines

External contributions and feedback are important to the development and future maintenance of FIMS and are welcome. This section provides guidelines and workflows for FIMS developers and collaborators on how to contribute to the project.

## Style Guide

The FIMS project uses style guides to ensure our code is consistent, easy to use (e.g. read, share, and verify), and ultimately easier to write. We use the [Google C++ Style Guide](https://google.github.io/styleguide/cppguide.html) and [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html). Google's R Style Guide is based off of the [tidyverse style guide](https://style.tidyverse.org/), with a few minor modifications to improve readability and portability.

## Naming Conventions

The FIMS implementation team has chosen to use `typename` instead of `class` when defining templates for consistency with the `TMB` package. While types may be defined in many ways, for consistency developers are asked to use `Type` instead of `T` to define `Type`s within FIMS.

## Coding Good Practices

Following good software development and coding practices simplifies collaboration, improves readability, and streamlines testing and review. The following are industry-accepted standards:

* Adhere to the [FIMS Project style guide](https://noaa-fims.github.io/collaborative_workflow/style-guide.html)
* Avoid rework - take the time to check for existing options (e.g. in-house, open source, etc.) before writing code
* Keep code as simple as possible
* Use meaningful variable names that are easy to understand and clearly represent the data they store
* Use descriptive titles and consistent conventions for class and function names
* Use consistent names for temporary variables that have the same kind of role
* Add clear and concise coding comments
* Use consistent formatting and indentation to improve readability and organization
* Group code into separate blocks for individual tasks
* Avoid hard-coded values to ensure portability of code
* Follow the DRY principle - "Don't Repeat Yourself" (or your code)
* Avoid deep nesting
* Limit line length (wrap ~72 characters)
* Capitalize SQL queries so they are readily distinguishable from table/column names
* Lint your code

## Roadmap to FIMS File Structure and Organization

### Files that go in `inst/include`

#### common

This folder includes files that are shared between the interface, the `TMB` objective function, and the mathematics and population dynamics components of the package.

#### interface

This includes the R interface files.

#### population dynamics

There are subfolders underneath this folder that correspond to the 
different components of the population dynamics model. Each of the 
modules will need a <module_name>.hpp file that only consists of 
`#include` statements for the files under the subfolders.

In the subfolder, there will need to be one file called 
<module_name>_base.hpp that defines the base class for the module type.
The base class should only need a constructor method and a number of
methods (e.g. `evaluate()`) that are not specific to the type of 
functions available under the subfolders but reused for all objects of 
that class type.

### Files that go in `src/`

#### FIMS.cpp

This is the `TMB` objective function.

## GitHub Collaborative Environment

Communication is managed via the [NOAA-FIMS
Github](https://github.com/NOAA-FIMS) organization.

* New features requests and bugs should be submitted as issues to the
[FIMS development repo](https://github.com/NOAA-FIMS/FIMS/issues). For
guidelines on submitting issues, see [Issue Tracking](#issue-tracking).
* [GitProjects](https://github.com/orgs/NOAA-FIMS/projects/1) TODO: add
description * [GitHub Teams](https://github.com/orgs/NOAA-FIMS/teams)
TODO: add description * All contributers, both internal and external,
are required to abide by the [Code of Conduct](#code-of-conduct)

### FIMS Branching Strategy

There are several [branching
strategies](https://reviewpad.com/blog/github-flow-trunk-based-development-and-code-reviews/)
available that will work within the Git environment and other version
control systems. However, it is important to find a strategy that works
well for both current and future contributors. Branching strategies
provide guidance for how, when, and why branches are created and named,
which also ties into necessary guidance surrounding issue tracking.

The FIMS Project uses a [Scaled Trunk Based
Development](https://trunkbaseddevelopment.com/) branching strategy to
make tasks easier without compromising quality.

![Scaled Trunk Based Development; image credit:
https://reviewpad.com/blog/github-flow-trunk-based-development-and-code-reviews/](https://reviewpad.com/wp-content/uploads/2021/06/Scaled-Trunk-Based-Development.svg)

This  strategy is required for continuous integration and facilitates
knowledge of steps that must be taken prior to, during, and after making
changes to the code, while still allowing anyone interested in the code
to read it at any time. Additionally, trunk-based development captures
the following needs without being overly complicated:

* Short-lived branches to minimize stale code and merge conflicts * Fast
release times, especially for bug fixes * Ability to release bug fixes
without new features

### Branch Protection

Branch protection allows for searching branch names with `grep`
functionality to apply merging rules (i.e., protection). This will be
helpful to protect the main/trunk branch such that pull requests cannot
be merged in prior to passing various checks or by individuals without
the authority to do so.

### GitHub cloning and branching

For contributors with write access to the FIMS repo, changes should be made on a feature branch after cloning the repo. The FIMS repo can be cloned to a local machine by using on the command line:

```bash
git clone https://github.com/NOAA-FIMS/FIMS.git
```

### Outside collaborators and forks

Outside collaborators without write access to the FIMS repos will be required to fork the repository, make changes, and submit a pull request. Forks are discouraged for every-day development because it becomes difficult to keep track of all of the forks. Thus, it will be important for those working on forks to be active in the issue tracker in the main repository prior to working on their fork — just like any member of the organization would do if they were working within the organization. Knowledge of future projects, ideas, concerns, etc. should always be documented in an issue before the code is altered.

Pull requests from forks will be reviewed the same as a pull request submitted from a branch. Users will need to conform to the same standards and all contributions must pass the standard tests as well as have tests that check the new feature.

To fork and then clone a repository, follow the [Github Documentation for forking a repo](https://docs.github.com/en/get-started/quickstart/fork-a-repo#forking-a-repository). Once cloned, changes can be made on a feature branch. When ready to submit changes follow the [Github Documentation on creating a pull request from a fork](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request-from-a-fork)

## Issue Tracking

Use of the GitHub issue tracker is key to keeping everyone informed and prioritizing key tasks. All future projects, ideas, concerns, development,
etc. must be documented in an issue before the code is altered. Issues should be filed and tagged prior to any code changes whether the change
pertains to a bug or the development of a feature. At a minimum, all issues will be labeled with a future version number. Bugs with immediate fixes will
be assigned to the current version number augmented for a hot fix and development will be based on code in the trunk. All other issues will be
assigned to a future version and development will be based on version branches. That is, changes to the code for version 3.3 cannot start until there is a branch
for version 3.3. This will minimize stale code and large merge conflicts.

## Reporting Bugs

This section guides you through submitting a bug report for any toolbox tool. Following these guidelines helps maintainers and the community understand your report, reproduce the behavior, and find related reports.

#### Before Submitting A Bug Report

* **Check if it is related to version.** We recommend using `sessionInfo()` within your `R` console and submitting the results in your bug report. Also, please check your R version against the required R version in the DESCRIPTION file and update if needed to see if that fixes the issue.
* **Perform a cursory search of issues** to see if the problem has already been reported. If it has **and the issue is still open**, add a comment to the existing issue instead of opening a new one. If it has **and the issue is closed**, open a new issue and include a link to the original issue in the body of your new one.

#### How Do I Submit A (Good) Bug Report?

Bugs are tracked as [GitHub issues](https://guides.github.com/features/issues/). Create an issue on the toolbox Github repository and provide the following information by following the steps outlined in the [reprex package](https://reprex.tidyverse.org/articles/reprex-dos-and-donts.html). Explain the problem and include additional details to help maintainers reproduce the problem using the Bug Report issue [template](https://github.com/NOAA-FIMS/FIMS/issues/new/choose).

Provide more context by answering these questions:

* **Did the problem start happening recently** (e.g. after updating to a new version of R) or was this always a problem?
* If the problem started happening recently, **can you reproduce the problem in an older version of R?** What's the most recent version in which the problem doesn't happen? 
* **Can you reliably reproduce the issue?** If not, provide details about how often the problem happens and under which conditions it normally happens.
* If the problem is related to working with files (e.g. reading in data files), **does the problem happen for all files and projects or only some?** Does the problem happen only when working with local or remote files (e.g. on network drives), with files of a specific type (e.g. only JavaScript or Python files), with large files or files with very long lines, or with files in a specific encoding? Is there anything else special about the files you are using?

Include details about your configuration and environment:

* **Which version of the tool are you using?** 
* **What's the name and version of the OS you're using?**
* **Which packages do you have installed?** You can get that list by running `sessionInfo()`.


## Suggesting Features

This section guides you through submitting an feature suggestion for toolbox packages, including completely new features and minor improvements to existing functionality. Following these guidelines helps maintainers and the community understand your suggestion and find related suggestions.

Before creating enhancement suggestions, please check the issues list as you might find out that you don't need to create one. When you are creating an enhancement suggestion, please include an "enhancement" tag in the issues.

#### Before Submitting A Feature Suggestion

* **Check you have the latest version of the package.**
* **Check if the development branch has that enhancement in the works.**
* **Perform a cursory search of the issues and enhancement tags** to see if the enhancement has already been suggested. If it has, add a comment to the existing issue instead of opening a new one.

#### How Do I Submit A (Good) Feature Suggestion?

Feature suggestions are tracked as [GitHub issues](https://guides.github.com/features/issues/). Create an issue on the repository and use the Feature Request issue [template](https://github.com/NOAA-FIMS/FIMS/issues/new/choose).

### Issue Labels

Utilize labels on issues:

* To describe the kind of work to be done: bug, enhancement, task, discussion, question, suitable for beginners
* To indicate the state of the issue: urgent, current, next, eventually, won't fix, duplicate

### Issue Templates

Templates are available and stored within each repository to guide users through the process of submitting a new issue.
Example templates for issues can be found on
[GitHub Docs](http://www.docs.github.com/en/communities/using-templates-to-encourage-useful-issues-and-pull-requests).
Use these references and existing templates stored in .github/ISSUE_TEMPLATE for reference when creating a new template.

## Branch Workflow

This section details the workflow to create a branch in order to contribute to FIMS.

### Branching Good Practices

The following suggestions will help ensure optimal performance of the trunk-based branching strategy:

1. Branches and commits should be kept small (e.g. a couple commits, a few lines of code) to allow for rapid merges and deployments.
2. Use [feature flags](https://martinfowler.com/articles/feature-toggles.html) to wrap new changes in an inactive code path for later activation (rather than creating a separate repository feature branch).
3. Delete branches after it is merged to the trunk; avoid repositories with a large number of "active" branches.
4. Merge branches to the trunk frequently (e.g. **at least** every few days; tag as a release commit) to avoid merge conflicts.
5. Use caching layers where appropriate to optimize build and test execution times.

### Branch Naming Conventions

Example: **feat-issue-3-R-package-skeleton**

1. When creating a branch, start the branch name with a group word:

  - **bug**: bug fix

  - **feat**: new feature

2. Use issue tracker ID in branch name

3. Include brief description

3. Use a hyphen as separators

### git workflow

1. Use the following commands to create a branch:

```{r eval = FALSE}
$ git checkout -b <branchname> main           #creates a local branch
$ git push origin <branchname>                #pushes branch back to gitHub
```

2. Periodically merge changes from main into branch
```{r eval = FALSE}
$ git merge main                              #merges changes from main into branch
```

3. While editing code, commit regularly following [commit messages](#commit-messages) guidelines
```{r eval = FALSE}
$ git add <filename>                          #stages file for commit
$ git commit -m"Commit Message"               #commits changes
```

4. To push changes to gitHub, first set the upstream location:
```{r eval = FALSE}
$ git push --set-upstream origin <branchname> #pushes change to feature branch on gitHub
```
After which, changes can be pushed as:
```{r eval = FALSE}
$ git push                 #pushes change to feature branch on gitHub
```

4. When finished, create a pull request to the main branch following
[pull request](#pull-request) guidelines


## Code Development
Code is written following the [Style Guide](#style-guide),
[FIMS Naming Conventions](#naming conventions), and
[Coding Good Practices](#coding-good-practices)

## Commit Messages

FIMS Project contributors should provide clear, descriptive commit
messages to communicate to collaborators details about changes that have
occurred and improve team efficiency. Good commit messages follow the
following practices:

* Include a short summary of the change for the subject/title (<50 characters)
* Include a blank line in between the 'subject' and 'body'
* Specify the type of commit:

          * fix: bug fix
          * feat: new feature
          * test: testing
          * docs: documentation
          * chore: regular code maintenance (e.g. updating dependencies)
          * refactor: refactoring codebase
          * style: changes that do not affect the meaning of the code; instead address code styling/formmatting
          * perf: performance improvements
          * revert: reverts a previous commit
          * build: changes that affect the build system

* If the commit addresses an issue, indicate the issue# in the title
* Provide a brief explanatory description of the change, addressing what and why was changed
* Wrap to ~72 characters
* Write in the imperative (e.g. "Fix bug", not "Fixed bug")
* If necessary, separate paragraphs by blank lines
* Utilize `BREAKING CHANGE: <description>` to provide expanation or further context about the issue being addressed.
* If the commit closes an issue, include a footer to note that (i.e. "Closes #19")


## Merge Conflicts

### What is a merge conflict?

A merge conflict happens when changes have occured to the same piece of code on the two branches being merged. This means Git cannot automatically determine which version of the change should be kept. Most merge conflicts are small and easy to figure out. See the [Github Documentation on merge conflicts](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/about-merge-conflicts) for more information.

### How to prevent merge conflicts

- Merge in small changes often rather than making many changes on a branch that is kept separate from the main branch for a long time.
- Avoid refactoring the same piece of code in different ways on separate branches.
- Avoid working in the same files on separate branches.

### How to resolve merge conflicts

Merge conflicts can be resolved [on Github](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/resolving-a-merge-conflict-using-the-command-lineithub.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/about-merge-conflicts) or [locally using Git](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/resolving-a-merge-conflict-using-the-command-line).

An additional helpful resource is [this guide to merge conflicts](https://github.com/TatianaTylosky/solving-and-preventing-merge-conflicts/blob/master/guide.md).

## Pull Requests

Once development of a module is complete, the contributor must initiate
a pull request. Github will automatically start an independent
[review](#code-review) process before the branch can be merged back into
the main development branch. Pull requests are used to identify changes
pushed to development branches. Open pull requests allow the FIMS
Development Team to discuss and review the changes, as well as add
follow-up commits before merging to the main branch. As noted in the
[branching stratgegy](#branching-strategy) section, branches, commits,
and pull requests should be kept small to enable rapid review and reduce
the chance of merge conflicts. Any pull requests for the FIMS Project
must be fully tested and reviewed before being merged into the main
branch.
Use the [pull request template](https://github.com/NOAA-FIMS/FIMS/blob/main/.github/PULL_REQUEST_TEMPLATE/pull_request_template.md)
to create pull requests. Pull requests without this template attached
will not be approved.

## Code Review

Code review ensures health and continuous improvement of the FIMS
codebase, while simultaneously helping FIMS developers become familiar
with the codebase and ensure there is a diverse team of knolwedgable
collaborators to support the continued development and maintenance of
FIMS. CI/CD requires rapid review of all new/modified code, so processes
must be in place to support this pace. FIMS code review will utilize
tools available via [GitHub](https://github.com/features/code-review),
which allows reviewers to analyze code changes, provide inline comments,
and view change histories.

[The Google code review developer guide](https://google.github.io/eng-practices/review/) provides a useful set of guidelines for both reviewers and code authors.

### Assigning Reviewers

Reviewers for the FIMS Project may be assigned in two different ways:

1. A specific member of the FIMS Development Team is requested to review
a pull request, based on their specific expertise. 
2. Code review assignments are automatically assigned using the
[GitHub load balance routing algorithm](https://docs.github.com/en/organizations/organizing-members-into-teams/managing-code-review-settings-for-your-team);
this approach tries to ensure that each team member reviews an equal
number of pull request in any 30 day period.

Team members should keep their status in Github current
(see ["Setting a
status"](https://docs.github.com/en/account-and-profile/setting-up-and-managing-your-github-profile/customizing-your-profile/personalizing-your-profile#setting-a-status)
for more information). Reviews will not be auto-assigned to "Busy" team
members.

If a review has been assigned to you and you don't feel like you have
the expertise to address it properly, please respond directly to the
code owner immediately so a different reviewer can be found promptly.

### Automated Testing

Automated testing provides an initial layer of quality assurance and
lets reviewers know that the code meets certain standards. For more on
FIMS testing, see [Testing](#testing) and [GitHub Actions](#github-actions).

### Review Checklist

While automated testing can assure the code structure and logic pass
quality checks, human reviewers are required to evaluate things like
functionality, readability, etc. Every pull request is accompanied by an 
automatically generated checklist of major considerations for code reviews;
additional guidance is provided below for reviewers to evaluate when 
providing feedback on code:

* **Design** (Is the code in the proper location? Are files organized
intuitively? Are components divided up in a sensible way? Does the pull 
request include an appropriate number of changes, or would the code 
changes be better broken into more focused parts? Is the code focused on
only requirements within the current scope? Does the code follow object-
oriented design principles? Will changes be easy to maintain? Is the code 
more complex than it needs to be?)

* **Functionality** (Does the code function as it is expected to? Are
changes, including to the user interface (if applicable), good for users?
Does parallel computing remain functional? How will the change impact 
other parts of the system? Are there any unhandled edge cases? Are there 
other code improvements possible?)

* **Testing** (Does the code have appropriate [unit tests](#testing)? Are tests well-
designed? Have dependencies been appropriately tested? Does automated 
testing cover the code exchange adequately? Could the test structure be improved?)

* **Readability** (Is the code and data flow easy to understand? Are there 
any parts of the code that are confusing or commented out? Are names clear? 
Does the code include any errors, repeats, or incomplete sections? Does the code 
adhere to the [FIMS Style Guide](#contributor-guide)?)

* **Documentation** (Are there clearl and useful comments available to why
the code has been implemented as it has been? Is the code appropriately documented
(doxygen and roxygen)? Is the README file complete, current, and adequately 
describe project/changes?)

* **Security** (Does using this code open the software to possible security 
violations or vulnerabilities?)

* **Performance** (Are there ways to improve on the code's performance?
Is there any complex logic that could be simplified? Could any of the code be 
replaced with built-in functions? Will this change have any impacts on system 
performance? Is there any debugging code that could be removed? Are there any 
optimizations that could be removed and still maintain system performance?)

### Review Good Practices

Good reviews require good review habits. Try to follow these suggestions:

* Review in short sessions (< 60 minutes) to maintain focus and
attention to detail
* Don't try to review more than 400 lines of code in a single session
* Provide constructive and supportive feedback
* Ask open-ended questions and offer alternatives or possible workarounds
* Avoid strong/opinionated statements
* Applaud good solutions
* Don't say "you"
* Be clear about which questions/comments are non-blocking or
unimportant; likewise, be explicit when approving a change or requesting
follow-up
* Aim to minimize the number of nitpicks (if there are a lot, suggest a
team-level resolution)
* Use the [FIMS Style Guide](https://noaa-fims.github.io/collaborative_workflow/style-guide.html)
to settle any style arguments

## Clean up local branches
If a code reviewer approves the pull request, FIMS workflow managers
will merge the feature/bug branch back into the main repository and
delete the branch. At this stage, the contributor should also delete the
branch from the local repository using the following commands:

```{r, eval = FALSE}
$ git checkout main           //switches back to main branch
$ git branch -d <branchname>  //deletes branch from local repository
```
## GitHub Actions

FIMS uses [GitHub Actions](https://docs.GitHub.com/en/actions) to automate
routine tasks. These tasks include:

1. Backup checks for developers
2. Routine GitHub workflow tasks (not important for developers to monitor)

Currently, the GitHub Actions in the FIMS repository include:

| GitHub Action Name  | Description | Type | Runs a Check on PRs? | Runs on: |
| ------------------- | ------------------------------ | ------------ | ------ | ----------|
| call-r-cmd-check    | Runs R CMD Check               | Backup Check | Yes | Push to any branch |
| run-clang-tidy      | Checks for C++ code            | Backup Check | Yes | Push to any branch |
| run-googletest      | Runs the google C++ unit tests | Backup Check | Yes | Push to any branch |
| run-doxygen         | Builds the C++ documentation   | Backup Check | No | Push to main branch |
| run-clang-format    | Styles C++ code             | Routine workflow task | No | Push to main branch |
| call-doc-and-style-r| documents and styles R code | Routine workflow task | No  | Push to main branch |
| pr-checklist | Generates a checklist as a comment for reviewers on PRs | Routine workflow task | No | Opening a PR |

[YAML files in a subdirectory of the FIMS repository](https://github.com/NOAA-FIMS/FIMS/tree/main/.github/workflows) specify the setup for the GitHub Actions. Some of
the actions depend on [reusable
workflows](https://docs.GitHub.com/en/actions/using-workflows/reusing-workflows)
available in [{ghactions4r}](https://nmfs-fish-tools.github.io/ghactions4r/).

Runs of the GitHub Actions can be viewed by navigating to the [Actions tab of
the FIMS repository](https://github.com/NOAA-FIMS/FIMS/actions). The status of 
GitHub Action runs can also be viewed on pull requests or next to commits
throughout the FIMS repository.

### Details on Backup Checks

Developers must make sure that the checks on their pull requests pass, as
typically changes will not be merged into the main branch until all GitHub
Actions are passing (the exception is if there are known reasons for the GitHub
Actions to fail that are not related to the pull request). Other
responsibilities of developers are listed in [the Code Development
section](#code-development).

Additional details about the backup check GitHub Actions:

- **call-r-cmd-check** runs R CMD Check on the FIMS package using the current
version of R. Three runs occur simultaneously, on three operating systems:
Windows, Linux (Ubuntu), and OSX. R CMD Check ensures that the FIMS package can
be downloaded without error. An error means that the package cannot be
downloaded successfully on the operating system for the run that failed.
Developers should investigate the failing runs and make fixes. To replicate the
GitHub Actions workflow locally, use `devtools::check()`
- **run-clang-tidy** runs checks while compiling the C++ code. If this run
  fails, fixes need to be made to the C++ code to address the issue identified.
- **run-googletest** Runs the GoogleTest C++ unit tests and benchmarking. If
 this run fails, then fixes need to be made to the C++ code and/or the
 GoogleTest C++ unit tests. To replicate this GitHub Actions workflow locally,
 follow instructions in the [testing section](#testing).

### Debugging Broken Runs

GitHub Actions can fail for many reasons, so debugging is necessary to find the
cause of the failing run. Some steps that can help with debugging are:

- Ask for help as needed! Some members of the FIMS team who have experience
debugging GitHub Actions are Bai, Kathryn, and Ian.
- Investigate why the run failed by [looking in the
log](https://docs.GitHub.com/en/actions/quickstart#viewing-your-workflow-results).
- Try to replicate the problem locally. For example, if the call-r-cmd-check run
  fails during the testthat tests, try running the testthat tests locally (e.g.,
using `devtools::test()`).
- If the problem can be replicated, try to fix locally by fixing one test or
issue at a time. Then push the changes up to GitHub and monitor the new Github
Action run.
- If the problem cannot be replicated locally, it could be a operating specific
issue; for example, if using Windows locally, it may be an issue specific to Mac
or Linux. 
- Sometimes, runs may fail because a particular dependency wasn’t available at
the exact point in time need for the run (e.g., maybe R didn’t install because
the R executable couldn’t be downloaded); if that is the case, wait a few hours
to a day and try to rerun. If it continues to fail for more than a day, a change
in the GitHub Action YAML file may be needed.

### How do I request a new Github Action workflow?

Routine actions and checks should be captured in a GitHub Action workflow in
order to improve efficiency of the development process and/or improve automated
checks on the FIMS codebase. New GitHub Action workflows can be requested by
opening an [issue in the FIMS
repository](https://github.com/NOAA-FIMS/FIMS/issues).
