# Contributor Guidelines

External contributions and feedback are important to the development and future maintenance of FIMS and are welcome. This section provides guidelines and workflows for FIMS developers and collaborators on how to contribute to the project.

## Style Guide

The FIMS project uses style guides to ensure our code is consistent, easy to use (e.g. read, share, and verify), and ultimately easier to write. We use the [Google C++ Style Guide](https://google.github.io/styleguide/cppguide.html) and [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html). Google's R Style Guide is based off of the [tidyverse style guide](https://style.tidyverse.org/), with a few minor modifications to improve readability and portability.

## Naming Conventions
TODO: add section with link to naming conventions used in FIMS

## Coding Good Practices

Following good software development and coding practices simplifies collaboration, improves readability, and streamlines testing and review. The following are industry-accepted standards:

* Adhere to the [FIMS Project style guide](https://noaa-fims.github.io/collaborative_workflow/style-guide.html)
* Avoid rework - take the time to check for existing options (e.g. in-house, open source, etc.) before writing code
* Keep code as simple as possible
* Use meaningful variable names that are easy to understand and clearly represent the data they store
* Use descriptive titles and consistent conventions for class and function names
* Use consistent names for temporary variables that have the same kind of role
* Add clear and concise coding comments
* Use consistent formatting and indentation to improve readability and organization
* Group code into separate blocks for individual tasks
* Avoid hard-coded values to ensure portability of code
* Follow the DRY principle - "Don't Repeat Yourself" (or your code)
* Avoid deep nesting
* Limit line length (wrap ~72 characters)
* Capitalize SQL queries so they are readily distinguishable from table/column names
* Lint your code

## GitHub forking and cloning

For contributors with write access to the FIMS repo, changes should be made on a feature branch after cloning the repo. The FIMS repo can be cloned to a local machine by using on the command line:

```bash
git clone https://github.com/NOAA-FIMS/FIMS.git
```

For contributors without write access to the FIMS repo, the repository must be forked to contribute. To fork and then clone a repository instead, follow the [Github Documentation for forking a repo](https://docs.github.com/en/get-started/quickstart/fork-a-repo#forking-a-repository). Once cloned, changes can be made on a feature branch. When ready to submit changes follow the [Github Documentation on creating a pull request from a fork](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request-from-a-fork)

## Issue Tracking

Use of the GitHub issue tracker is key to keeping everyone informed and prioritizing key tasks. All future projects, ideas, concerns, development,
etc. must be documented in an issue before the code is altered. Issues should be filed and tagged prior to any code changes whether the change
pertains to a bug or the development of a feature. At a minimum, all issues will be labeled with a future version number. Bugs with immediate fixes will
be assigned to the current version number augmented for a hot fix and development will be based on code in the trunk. All other issues will be
assigned to a future version and development will be based on version branches. That is, changes to the code for version 3.3 cannot start until there is a branch
for version 3.3. This will minimize stale code and large merge conflicts.

### Issue Labels

Utilize labels on issues:

* To describe the kind of work to be done: bug, enhancement, task, discussion, question, suitable for beginners
* To indicate the state of the issue: urgent, current, next, eventually, won't fix, duplicate

### Issue Templates

Templates are available and stored within each repository to guide users through the process of submitting a new issue.
Example templates for issues can be found on
[GitHub Docs](http://www.docs.github.com/en/communities/using-templates-to-encourage-useful-issues-and-pull-requests).
Use these references and existing templates stored in .github/ISSUE_TEMPLATE for reference when creating a new template.

## Branch Workflow

This section details the workflow to create a branch in order to contribute to FIMS.

### Branching Good Practices

The following suggestions will help ensure optimal performance of the trunk-based branching strategy:

1. Branches and commits should be kept small (e.g. a couple commits, a few lines of code) to allow for rapid merges and deployments.
2. Use [feature flags](https://martinfowler.com/articles/feature-toggles.html) to wrap new changes in an inactive code path for later activation (rather than creating a separate repository feature branch).
3. Delete branches after it is merged to the trunk; avoid repositories with a large number of "active" branches.
4. Merge branches to the trunk frequently (e.g. **at least** every few days; tag as a release commit) to avoid merge conflicts.
5. Use caching layers where appropriate to optimize build and test execution times.

### Branch Naming Conventions

Example: **feat-issue-3-R-package-skeleton**

1. When creating a branch, start the branch name with a group word:

  - **bug**: bug fix

  - **feat**: new feature

2. Use issue tracker ID in branch name

3. Include brief description

3. Use a hyphen as separators

### git workflow

1. Use the following commands to create a branch:

```{r eval = FALSE}
$ git checkout -b <branchname> main           #creates a local branch
$ git push origin <branchname>                #pushes branch back to gitHub
```

2. Periodically merge changes from main into branch
```{r eval = FALSE}
$ git merge main                              #merges changes from main into branch
```

3. While editing code, commit regularly following [commit messages](#commit-messages) guidelines
```{r eval = FALSE}
$ git add <filename>                          #stages file for commit
$ git commit -m"Commit Message"               #commits changes
```

4. To push changes to gitHub, first set the upstream location:
```{r eval = FALSE}
$ git push --set-upstream origin <branchname> #pushes change to feature branch on gitHub
```
After which, changes can be pushed as:
```{r eval = FALSE}
$ git push                 #pushes change to feature branch on gitHub
```

4. When finished, create a pull request to the main branch following
[pull request](#pull-request) guidelines


## Code Development
Code is written following the [Style Guide](#style-guide),
[FIMS Naming Conventions](#naming conventions), and
[Coding Good Practices](#coding-good-practices)

### C++ contributor

In addition to writing a C++ module, contributors are responsible for
the following:

* Creating documentation for the module and building the
[documentation](#documentation) in `doxygen` to ensure it is error-free.
* Implementing the suite of required test cases in
[Google Test](#testing) for the module.
* Ensuring the module runs through the FIMS Github actions.

### R Contributor

In addition to writing an R script or function, contributors are
responsible for the following:

* Writing Roxygen documentation for any R functions
* Writing `testthat()` test cases for any R functionality
* Ensuring the code passes the `R CMD CHECK`, `styler`, and any other
automated checks.

## Commit Messages

FIMS Project contributors should provide clear, descriptive commit
messages to communicate to collaborators details about changes that have
occurred and improve team efficiency. Good commit messages follow the
following practices:

* Include a short summary of the change for the subject/title (<50 characters)
* Include a blank line in between the 'subject' and 'body'
* Specify the type of commit:

          * fix: bug fix
          * feat: new feature
          * test: testing
          * docs: documentation
          * chore: regular code maintenance (e.g. updating dependencies)
          * refactor: refactoring codebase
          * style: changes that do not affect the meaning of the code; instead address code styling/formmatting
          * perf: performance improvements
          * revert: reverts a previous commit
          * build: changes that affect the build system

* If the commit addresses an issue, indicate the issue# in the title
* Provide a brief explanatory description of the change, addressing what and why was changed
* Wrap to ~72 characters
* Write in the imperative (e.g. "Fix bug", not "Fixed bug")
* If necessary, separate paragraphs by blank lines
* Utilize `BREAKING CHANGE: <description>` to provide expanation or further context about the issue being addressed.
* If the commit closes an issue, include a footer to note that (i.e. "Closes #19")


## Merge Conflicts

### What is a merge conflict?

A merge conflict happens when changes have occured to the same piece of code on the two branches being merged. This means Git cannot automatically determine which version of the change should be kept. Most merge conflicts are small and easy to figure out. See the [Github Documentation on merge conflicts](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/about-merge-conflicts) for more information.

### How to prevent merge conflicts

- Merge in small changes often rather than making many changes on a branch that is kept separate from the main branch for a long time.
- Avoid refactoring the same piece of code in different ways on separate branches.
- Avoid working in the same files on separate branches.

### How to resolve merge conflicts

Merge conflicts can be resolved [on Github](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/resolving-a-merge-conflict-using-the-command-lineithub.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/about-merge-conflicts) or [locally using Git](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/resolving-a-merge-conflict-using-the-command-line).

An additional helpful resource is [this guide to merge conflicts](https://github.com/TatianaTylosky/solving-and-preventing-merge-conflicts/blob/master/guide.md).

## Pull Requests

Once development of a module is complete, the contributor must initiate
a pull request. Github will automatically start an independent
[review](#code-review) process before the branch can be merged back into
the main development branch. Pull requests are used to identify changes
pushed to development branches. Open pull requests allow the FIMS
Development Team to discuss and review the changes, as well as add
follow-up commits before merging to the main branch. As noted in the
[branching stratgegy](#branching-strategy) section, branches, commits,
and pull requests should be kept small to enable rapid review and reduce
the chance of merge conflicts. Any pull requests for the FIMS Project
must be fully tested and reviewed before being merged into the main
branch.
Use the [pull request template](https://github.com/NOAA-FIMS/FIMS/blob/main/.github/PULL_REQUEST_TEMPLATE/pull_request_template.md)
to create pull requests. Pull requests without this template attached
will not be approved.

## Code Review

Code review ensures health and continuous improvement of the FIMS
codebase, while simultaneously helping FIMS developers become familiar
with the codebase and ensure there is a diverse team of knolwedgable
collaborators to support the continued development and maintenance of
FIMS. CI/CD requires rapid review of all new/modified code, so processes
must be in place to support this pace. FIMS code review will utilize
tools available via [GitHub](https://github.com/features/code-review),
which allows reviewers to analyze code changes, provide inline comments,
and view change histories.

[The Google code review developer guide](https://google.github.io/eng-practices/review/) provides a useful set of guidelines for both reviewers and code authors.

### Assigning Reviewers

Reviewers for the FIMS Project may be assigned in two different ways:

1. A specific member of the FIMS Development Team is requested to review
a pull request, based on their specific expertise. 2. Code review
assignments are automatically assigned using the
[GitHub load balance routing algorithm](https://docs.github.com/en/organizations/organizing-members-into-teams/managing-code-review-settings-for-your-team);
this approach tries to ensure that each team member reviews an equal
number of pull request in any 30 day period.

Team members should keep their status in Github current
(see ["Setting a
status"](https://docs.github.com/en/account-and-profile/setting-up-and-managing-your-github-profile/customizing-your-profile/personalizing-your-profile#setting-a-status)
for more information). Reviews will not be auto-assigned to "Busy" team
members.

If a review has been assigned to you and you don't feel like you have
the expertise to address it properly, please respond directly to the
code owner immediately so a different reviewer can be found promptly.

### Automated Testing

Automated testing provides an initial layer of quality assurance and
lets reviewers know that the code meets certain standards. For more on
FIMS testing, see [Testing](#testing).

### Review Checklist

While automated testing can assure the code structure and logic pass
quality checks, human reviewers are required to evaluate things like
functionality, readability, etc. Reviewers should evaluate the code
critically and provide comment/feedback on the following items:

**Readability**

* Is the code easy to understand?
* Are there any parts of the code that are confusing?
* Is the data flow easy to understand?
* Is there any code commented out?
* Does the code include any unclear names
* Does the code include any errors, repeats, or incomplete sections?

**Functionality**

* Does the code function as it is expected to?
* How will the change impact other parts of the system?
* Are there any unhandled edge cases?
* Are there other code improvements possible?

**Design**

* Are files organized intuitively?
* Are components divided up in a sensible way?
* Does the review include too many changes? Would the code change better
be broken into more focused parts?
* Will the change be easy to maintain?
* Does the code follow object-oriented design principles?
* Is the code in the proper location?

**Security**

* Does using this code open the software to possible security violations
or vulnerabilities?
* Is the correct encryption used?

**Performance**

* Are there ways to improve on the code's performance?
* Is there any complex logic that could be simplified?
* Could any of the code be replaced with built-in functions?
* Will this change have any impacts on system performance?
* Is there any debugging code that could be removed?
* Are there any optimizations that could be removed and still maintain
system performance?

**Documentation**

* Are there comments available to explain the code?
* Is the README file complete and current? Does it adequately describe
the project/changes?

**Testing**

* Is the code testable?
* Is the automated testing adequate?
* Have dependencies been appropriately tested?
* Does automated testing cover the code exchange adequately?
* Could the test structure be improved?

### Review Good Practices

Good reviews require good review habits. Try to follow these suggestions:

* Review in short sessions (< 60 minutes) to maintain focus and
attention to detail
* Don't try to review more than 400 lines of code in a single session
* Provide constructive and supportive feedback
* Ask open-ended questions and offer alternatives or possible workarounds
* Avoid strong/opinionated statements
* Applaud good solutions
* Don't say "you"
* Be clear about which questions/comments are non-blocking or
unimportant; likewise, be explicit when approving a change or requesting
follow-up
* Aim to minimize the number of nitpicks (if there are a lot, suggest a
team-level resolution)
* Use the [FIMS Style Guide](https://noaa-fims.github.io/collaborative_workflow/style-guide.html)
to settle any style arguments

## Clean up local branches
If a code reviewer approves the pull request, FIMS workflow managers
will merge the feature/bug branch back into the main repository and
delete the branch. At this stage, the contributor should also delete the
branch from the local repository using the following commands:

```{r, eval = FALSE}
$ git checkout main           //switches back to main branch
$ git branch -d <branchname>  //deletes branch from local repository
```
